Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-04-30 doc/sat2023-rebuttal.txt

We wish to thank the reviewers for their helpful comments.  We will first respond to the most important items overall and then to ones by specific reviewers.

DETERMINISTIC vs. DECISION
We will be more careful about the distinction between these two flavors of DNNFs.  Our proof framework supports DETERMINISTIC, although our current proof generator relies on the inputs being in DECISION form.  We will clarify this.

PRIOR WORK
It was a major oversight on our part not to have cited the work by Capelli, Lagniez, and Marquis in AAAI-21.  This is, of course, very relevant.  Looking in detail at the conditions they check, it can be seen that we do much of the same work in our proof generator.  The main difference is that we then emit information as lines in a well-defined, formally-verified clausal proof framework.  As an example, they check that every path to a shared node in the decision-DNNF graph yields an equivalent simplified set of input clauses.  Our proof generator checks this too, but then uses this information to apply a lemma, encoded as a sequence of clausal assertions.  Our checker need not have any understanding of simplified input clauses or lemmas---it simply performs a sequence of RUP checks.

We have been unable to locate a copy of their checker online.  If we can find one, we'll do a comparison of runtimes and proof sizes.  D4 can emit a DRAT version of its conflict clauses.  At the very least, we will measure the sizes and checking times for these.

@@

GENERALITY
Although we present results only for results from D4, we have also successfully used it on the results from DSHARP.  We have found that DSHARP lags far behind D4 in terms of both time and the size of the generated result.  Our proof framework is sufficiently general to handle SDDs.  We will list that as future work.

Reviewer #2’s comment that “Basically the knowledge compiler could be a black-box” sparked us to consider whether we could deal with formulas that have been preprocessed using equivalence-preserving transformations.  We were able to modify the source code of D4 to enable its optional preprocessing.  (This had been disabled in the version on Github.) The structured approach does not work, but the monolithic approach works fine.  We have run the same benchmarks as we did for testing monolithic generation.  On average, performing the three supported preprocessing steps improves the overall runtime (D4+generator+checker), with a harmonic mean speedup of 2.7X.  We will present data on this.  We do not believe the CD4 program is able to generate end-to-end proofs of this form.

@@

REVIEWER#1

Your suggestions on how to make the presentation more clear are helpful.  The flow is not ideal, in part because we had to move so much information to the appendix to stay within the page limit.

### 2023-05-08 doc/mc-data.tex
Update statistics

\section*{Methodology}
\begin{enumerate}
   \item Retrieved 200 public model counting benchmarks from 2022 Model Counting Competition website
     \begin{center}
       \url{https://mccompetition.org/2022/mc_description.html}
     \end{center}
     \begin{itemize}
     \item 100 public instances from Track 1 (unweighted model counting)
     \item 100 public instances from Track 2 (weighted model counting)
     \end{itemize}
   \item Removed duplicates
     \begin{itemize}
    \item  20 instances were duplicated in the two tracks
     \item Reduces set of benchmarks to 180
     \end{itemize}
   \item Ran D4 on 180 instances with time limit of 4000 seconds
     \begin{itemize}
     \item Completed for 124 instances
     \end{itemize}
   \item Generated POG representations from dDNNF graphs
     \begin{itemize}
     \item Simplifications via constant propagation
       \begin{itemize}
       \item Number of POG nodes around $0{.}90\times$ number of dDNNF nodes
       \end{itemize}
     \item Used heuristics to identify root node, and then put in topological order with root node at end.
     \item Measure complexity by number of defining clauses $D$
       \begin{itemize}
       \item Product or sum node with degree $k$ requires $k+1$ clauses
       \item $D$ also equals sum of number of edges and number of nodes
       \item For 124 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
       \end{itemize}
     \end{itemize}
   \item Ran program {\sc d2p} to generate CPOG file for maximum of 10,000 seconds
     \begin{itemize}
     \item Completed for 108 instances
     \item Exceeded clause limit for 1 instance (Our programs represent a clause ID as a 32-bit signed value)
     \item Ran out of memory for 1 instance
     \item Timed out for 14 instances
     \item Was able to do one-sided verification for 9 of these.
     \end{itemize}
   \item For completed CPOG files, ran {\sc cpog-check} CPOG checker
     \begin{itemize}
     \item Tests all conditions for correct CPOG
     \item Most time spent performing RUP checks for clause addition and deletion
     \item Completed successfully for all 108 instances
     \item Time generally faster than {\sc d2p}. (min = 0.005, max = 1.4, harmonic mean = 0.10)
     \end{itemize}
\end{enumerate}

\section*{Time to Generate and Validate CPOG}

\begin{itemize}
\item Min 0.01, Average 1318.3, Max 13,144.8, Median 71.64
\item Ratio to time for D4: Min 0.50, Harmonic Mean 5.50, Max 457.69
\end{itemize}

@@

\input{figure-clauses-d4-verify}

\begin{itemize}
\item Implemented package to represent numbers of form $a \cdot 2^b \cdot 5^c$.
  \begin{itemize}
  \item For unweighted counts, first compute density and then scale by $2^n$
  \item For unweighted counts, perform exact decimal arithmetic
  \item The set of numbers of this form is not closed under a reciprocal operation, but the weights in the competion are either of the form
    $w(x) + w(\overline{x}) = 1.0$ or $w(x) = w(\overline{x}) = 1$.  The latter case sums to 2, with reciprocal $1/2$.
  \end{itemize}
\item Can implement having $a$ be integer of arbitrary size and $b$ and $c$ being 32-bit integers.
\item Weighted counts ranged up to 260,909 digits.
\end{itemize}

@@

The time performance is less dramatic, but still significant.
Comparing the sum of CPOG generation and checking times for each of
the four combinations of options, we get the following times relative
to the one with both optimizations enabled:
\begin{description}
\item[Without lemmas:] Min 0.85, Max 37.98, Harmonic mean 1.80
\item[Without literal grouping:] Min 0.83, Max 3.92, Harmonic mean 1.12
\item[Without either optimization:] Min 0.91, Max 41.97, Harmonic mean 2.19
\end{description}

Here is another way to summarize the impact on the combined time to generate and validate the CPOG file:
\begin{itemize}
\item Of the 80 benchmark problems evaluated, and comparing with both
  optimizations disabled, 40 had the time reduced by at least a factor of 2
  using lemmas, 7 by using literal merging, and 49 by enabling both
  optimizations.
\end{itemize}

@@

Ran all benchmarks for which generation + checking could complete in 1000 seconds.  Total = 82.  All completed in less than 1000 seconds with structured.
Compare both runtimes and total number of clauses for structured vs. monolithic:
\begin{itemize}
\item Structured complete 83.
\item Monolithic completed  70,timed out on 13.
\item Of the 70 completed, 23 were faster with monolithic, 37 were slower (min ratio = 0.015, max ratio = 9.66, hmean = 0.40)
\item Of the 70 completed, 60 had fewer total clauses with monolithic, 10 had more (min ratio = 0.07, max ratio = 1.09, hmean = 0.49)
\end{itemize}

@@

For comparison, we also ran the MICE tool chain for certifying model
counters~\cite{fichte:sat:2022}.  Figure~\ref{fig:mice} compares the
runtimes of our toolchain versus theirs.  The X-axis shows the sum of
the elapsed times for our generator, checker, and counter (since the MICE checker also provides a count), while the
Y-axis shows the sum of the elapsed times for \textsc{nnf2trace},
generating a file in their trace format from a d-DNNF graph generated
by D4, and for \textsc{sharptrace}, a checker for the trace format.
The plot shows data points for those benchmarks for at least one of
the two tools completed in less than 1000 seconds.

We encountered a significant number of errors with both
\textsc{nnf2trace} and \textsc{sharptrace}.  Based on the error
messages, these appear to be mostly due to features of some of the
CNF files or d-DNNF graphs they had not anticipated.  We provided a bug report to
the authors, but they were unable to fix them on short notice.

>>>>

\section*{Methodology}
\begin{enumerate}
   \item Retrieved 200 public model counting benchmarks from 2022 Model Counting Competition website
     \begin{center}
       \url{https://mccompetition.org/2022/mc_description.html}
     \end{center}
     \begin{itemize}
     \item 100 public instances from Track 1 (unweighted model counting)
     \item 100 public instances from Track 2 (weighted model counting)
     \end{itemize}
   \item Removed duplicates
     \begin{itemize}
    \item  20 instances were duplicated in the two tracks
     \item Reduces set of benchmarks to 180
     \end{itemize}
   \item Ran D4 on 180 instances with time limit of 4000 seconds
     \begin{itemize}
     \item Completed for 124 instances
     \end{itemize}
   \item Generated POG representations from dDNNF graphs
     \begin{itemize}
     \item Simplifications via constant propagation
       \begin{itemize}
       \item Number of POG nodes around $0{.}90\times$ number of dDNNF nodes
       \end{itemize}
     \item Used heuristics to identify root node, and then put in topological order with root node at end.
     \item Measure complexity by number of defining clauses $D$
       \begin{itemize}
       \item Product or sum node with degree $k$ requires $k+1$ clauses
       \item $D$ also equals sum of number of edges and number of nodes
       \item For 124 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
       \end{itemize}
     \end{itemize}
   \item Ran program {\sc d2p} to generate CPOG file for maximum of
     10,000 seconds.  Verifies file, generates unweighted count. If
     finds literal weights, also generates weighted count
     \begin{itemize}
     \item Completed for 108 instances
     \item Exceeded clause limit for 1 instance (Our programs represent a clause ID as a 32-bit signed value)
     \item Ran out of memory for 1 instance
     \item Timed out for 14 instances
     \item Was able to do one-sided verification for 9 of these.
     \end{itemize}
   \item For completed CPOG files, ran {\sc cpog-check} CPOG checker
     \begin{itemize}
     \item Tests all conditions for correct CPOG
     \item Most time spent performing RUP checks for clause addition and deletion
     \item Completed successfully for all 108 instances
     \item Time generally faster than {\sc d2p}. (min = 0.006, max = 1.38, harmonic mean = 0.107)
     \end{itemize}
\end{enumerate}

\begin{itemize}
\item Implemented C package to represent numbers of form $a \cdot 2^b \cdot 5^c$.
  \begin{itemize}
  \item For unweighted counts, first compute density and then scale by $2^n$
  \item For unweighted counts, perform exact decimal arithmetic
  \item The set of numbers of this form is not closed under a reciprocal operation, but the weights in the competion are either of the form
    $w(x) + w(\overline{x}) = 1.0$ or $w(x) = w(\overline{x}) = 1$.  The latter case sums to 2, with reciprocal $1/2$.
  \end{itemize}
\item Can implement having $a$ be integer of arbitrary size and $b$ and $c$ being 32-bit integers.
  \begin{itemize}
    \item Represent $a$ by ``digits,'' each ranging in value from $0$ to $10^9-1$.
    \item Fits in 32-bit number
    \item Easy scaling by 10
    \item Easy conversion to/from decimal string
  \end{itemize}
\item Weighted counts ranged up to 260,909 decimal digits.
\end{itemize}

\section*{Time to Generate, Validate, and Count CPOG}

\begin{itemize}
\item Min 0.012, Average 1,329.8, Max 13,198.3, Median 75.4
\item Ratio to time for D4: Min 0.51, Harmonic Mean 5.87, Max 465.8
\end{itemize}

@@

\input{figure-clauses-d4-result}

@@

The time performance is less dramatic, but still significant.
Comparing the sum of CPOG generation and checking times for each of
the four combinations of options, we get the following times relative
to the one with both optimizations enabled:
\begin{description}
\item[Without lemmas:] Min 0.89, Max 39.2, Harmonic mean 1.71
\item[Without literal grouping:] Min 0.80, Max 3.77, Harmonic mean 1.11
\item[Without either optimization:] Min 0.96, Max 43.69, Harmonic mean 2.20
\end{description}

Here is another way to summarize the impact on the combined time to generate and validate the CPOG file:
\begin{itemize}
\item Of the 80 benchmark problems evaluated, and comparing with both
  optimizations disabled, 40 had the time reduced by at least a factor of 2
  using lemmas, 6 by using literal merging, and 48 by enabling both
  optimizations.
\end{itemize}

@@

Ran all benchmarks for which generation + checking could complete in 1000 seconds.  Total = 83.  All completed in less than 1000 seconds with structured.
Compare both runtimes and total number of clauses for structured vs. monolithic:
\begin{itemize}
\item Structured complete 83.
\item Monolithic completed  70,timed out on 13.
\item Of the 70 completed, 23 were faster with monolithic, 47 were slower (min ratio = 0.013, max ratio = 9.60, hmean = 0.36)
\item Of the 70 completed, 60 had fewer total clauses with monolithic, 10 had more (min ratio = 0.07, max ratio = 1.09, hmean = 0.49)
\end{itemize}

\section{Preprocessing}

\input{figure-runtime-preprocess}
\input{figure-clauses-preprocess}

Use preprocessor in D4.  Enable its three preprocessing methods
(backbone, vivification, occElimination), all of which preserve
equivalence.  Cannot use structured mode for forward implication
proof, since modified clauses do not decompose in the way anticipated.

Consider same 83 benchmarks used to compare monolithic vs. structured mode.
Incorporate time to run D4 into evaluation, since differs according to
how D4 runs.

Runtime results:
\begin{itemize}
\item Monolithic without preprocessing completes 70  of them with an overall time less than 1000 seconds.
\item Monolithic with prepreprocessing completes 69 (the one due to too much time on preprocessing)
\item Of the 69, 53 run faster with preprocessing and 15 run slower (mininum ratio 0.04, maximum ratio 1.65, harmonic mean 0.38)
\end{itemize}

Clause results:
\begin{itemize}
\item Consider the 69 cases that completed for both with overall time less than 1000 seconds.
\item 31 had fewer clauses with preprocessing, 17 were identical, and 21 had more.
\item Min = 0.21, Max = 2.34, Harmonic mean = 0.85
\end{itemize}

@@

For comparison, we also ran the MICE tool chain for certifying model
counters~\cite{fichte:sat:2022}.  Figure~\ref{fig:mice} compares the
runtimes of our toolchain versus theirs.  The X-axis shows the sum of
the elapsed times for our generator and checker/counter, while the
Y-axis shows the sum of the elapsed times for \textsc{nnf2trace},
generating a file in their trace format from a d-DNNF graph generated
by D4, and for \textsc{sharptrace}, a checker and unweighted counter for the trace format
The plot shows data points for those benchmarks for at least one of
the two tools completed in less than 1000 seconds.
